Modelling with ANIMO:\\ between Fuzzy Logic and ODE

### 2015-05-01 Paper.tex
Updated with Janine's comments

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data which an experimental biologist
must deal with makes it nearly impossible to reason on
the dynamics of complex systems without relying on computational
support.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows to reason about the complex networks of signaling events occurring in living cells.
The user-friendly interface allows biology experts to exploit formal methods
without the need to become experts also in mathematical formalisms.
\parttitle{Results}
ANIMO models show a degree of precision that is intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models allow to replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic: for this reason, we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modeling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm for enhancing network
diagrams with an underlying formal description of both network components and their interactions.
Executable models typically contain a description of the direct interactions of network components,
requiring the modeler to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdeling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
 biologists without a thorough mathematical background. Such models are indispensable for formally
comparing experimental data to prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks, executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of biological network enriched with dynamic information.
ANIMO is based on Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node color on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

@@

Figure~\ref{fig:animo-spectrum} shows the position of ANIMO into a spectrum of modeling methods: justification for this placement will be given in the Results section.

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
Results obtained with ANIMO are comparable to results from widely used modeling
approaches. To demonstrate this, we present in Figure~\ref{fig:drosophila-model}
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}, based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how
ANIMO can be used to create reference models and help to reason on complex biological events.

@@

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-all}B). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~\ref{fig:large-model-hypotheses}, arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-complete}B). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature.
Transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}A).

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signaling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: updated version.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colors represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

In the Results section we also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct insight in the dynamic behavior of the network (see Fig.~\ref{fig:animo-spectrum}). Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions
that are involved in activation or inhibition.

In this study, we used a data set based on the experiments described in~\cite{pathway-compendium}. We used the
resulting experimental data, together with knowledge from curated databases~\cite{kegg,phosphosite}
to construct an executable model of the biological system.
In contrast with the two approaches described above, ANIMO is aimed at the construction of
more mechanistic models, mimicking biochemical interactions \emph{in silico}. This way of modeling
gives a different type of insight. In the process of model construction, we extended a
prior-knowledge network with time-dependent extracellular crosstalk that has been reported
previously~\cite{pathway-autocrine}. To come up with possible explanations for a disagreement
between the model and the experimental data, two additional layers of
crosstalk were introduced, at the signal transduction and transcriptional level. These modifications
improved the fit of the model to the data and can be interpreted as novel testable hypotheses.
Finally, we proposed new experiments that could be carried out to test these hypotheses, closing the empirical cycle.
Together, our model sheds more light on the intricate
entanglement between the TNF$\alpha$ and EGF pathways at multiple cellular levels.
But above all,  the model provides an excellent starting point for further investigation.
% Every new round in the empirical cycle will lift the understanding of the system to a higher level,
% leading to an incremental build-up of knowledge and an upward empirical spiral. Being intuitively
% accessible, ANIMO models facilitate sharing knowledge within and between groups and encourage collaborations.

\subsection*{Comparison between ANIMO and other modeling tools}\label{suppl:comparison-table}
Different formalisms are in use in the field of computational
modeling of biological systems, each with their specific characteristics.
Many of these formalisms have been implemented into
software tools to support modeling efforts. In order to compare
ANIMO with existing tools, we have selected a number of mathematical formalisms,
each connected to a supporting tool. With an emphasis on the modeling
process rather than the final model, we compared these tools on
the basis of the following parameters:

@@

All reactants modeled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions (see Methods Section). Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

In this paper we discussed the placement of ANIMO among other modeling paradigms and tools,
highlighting ANIMO's strong points. The aim is not to promote the use of ANIMO in any modeling effort, but to show where it best shines.

>>>>

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data an experimental biologist
must deal with makes it nearly impossible to reason on
the dynamics of complex systems without relying on computational
support.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signaling events occurring in living cells.
The user-friendly interface allows biology experts to exploit formal methods
without the need to become experts also in mathematical formalisms.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modeling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm for enhancing network
diagrams with an underlying formal description of both network components and their interactions.
Executable models typically contain descriptions of the direct interactions of network components,
requiring the modeler to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdeling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
 biologists without a thorough mathematical background. Such models are indispensable for formally
comparing experimental data to prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node color on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

@@

Figure~\ref{fig:animo-spectrum} shows the position of ANIMO in a spectrum of modeling methods: justification for this placement is given in the Results section.

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modeling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how
ANIMO can be used to create reference models and help to obtain insight into complex biological events.

@@

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-all}B). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~\ref{fig:large-model-hypotheses}, arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-complete}B). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature:
transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}A).

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signaling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colors represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

We also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct insight in the dynamic behavior of the network (see Fig.~\ref{fig:animo-spectrum}). Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points,
as measured in wet-lab experiments, were
not reproduced by the fuzzy logic model. Moreover, the fuzzy logic model gave no insight in the molecular interactions
that are involved in activation or inhibition.

Here we used a data set based on the wet-lab experiments described in~\cite{pathway-compendium}. We used the
resulting experimental data, together with knowledge from curated databases~\cite{kegg,phosphosite}
to construct an executable model of the biological system.
In contrast to the two approaches described above, ANIMO is aimed at the construction of
more mechanistic models, mimicking biochemical interactions \emph{in silico}. This way of modeling
gives a different type of insight. In the process of model construction, we extended a
prior-knowledge network with time-dependent extracellular crosstalk that has been reported
previously~\cite{pathway-autocrine}. To come up with possible explanations for a disagreement
between the model and the experimental data, two additional layers of
crosstalk were introduced, at the signal transduction and transcriptional level. These modifications
improved the fit of the model to the data and can be interpreted as novel testable hypotheses.
Finally, we proposed new experiments that could be carried out to test these hypotheses, closing the empirical cycle.
Together, our model sheds more light on the intricate
entanglement between the TNF$\alpha$ and EGF pathways at multiple cellular levels.
But above all,  the model provides an excellent starting point for further investigation.
% Every new round in the empirical cycle will lift the understanding of the system to a higher level,
% leading to an incremental build-up of knowledge and an upward empirical spiral. Being intuitively
% accessible, ANIMO models facilitate sharing knowledge within and between groups and encourage collaborations.

\subsection*{Comparison between ANIMO and other modeling tools}\label{suppl:comparison-table}
Different formalisms are in use in the field of computational
modeling of biological systems, each with their specific characteristics.
Many of these formalisms have been implemented into
software tools to support modeling efforts. To compare
ANIMO with existing tools, we have selected a number of mathematical formalisms,
each connected to a supporting tool. With an emphasis on the modeling
process rather than the final model, we compared these tools on
the basis of the following parameters:

@@

All reactants modeled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions (see Methods Section). Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the circadian clock
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

In this paper we discussed the placement of ANIMO among other modeling paradigms and tools,
highlighting ANIMO's strong points.
